---
title: Using nimbleQuad
author: "Paul van Dam-Bates and Chris Paciorek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  An overview of functioning in nimbleQuad for finding the MLE and posterior approximations.
vignette: >
  %\VignetteIndexEntry{Using nimbleQuad}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center",
  # eval = TRUE ## uncomment this to build quickly without running code.
)
```

## Introduction

The `nimbleQuad` package is an alternative to Markvo chain Monte Carlo algorithms in NIMBLE. It has two main components:

1. Adaptive Gauss-Hermite Quadrature (AGHQ) for marginalizing over latent nodes (e.g. random effects) to preform maximum likelihood estimation (`buildLaplace` if `nQuad = 1`) and `buldAGHQ` otherwise.
2. Quadrature methods for posterior approximation using Laplace for the inner marginalization of the latent nodes, and general quadrature methods for outer marginalization  (`buildNestedApprox`).

In this document we will introduce some of the workflow and show the custom control available when using `buildNestedApprox` for posterior approximations.

```{r, message = FALSE}
library(nimbleQuad)
```

### What is nimble?

NIMBLE is a system for writing hierarchical statistical models and algorithms.  It is distributed as an R package [nimble](https://CRAN.R-project.org/package=nimble).  NIMBLE stands for "Numerical Inference for statistical Models using Bayesian and Likelihood Estimation".  NIMBLE includes:

1. A dialect of the BUGS model language that is extensible.  NIMBLE uses almost the same model code as WinBUGS, OpenBUGS, and JAGS.  Being "extensible" means that it is possible to write new functions and distributions and use them in your models.

2. An algorithm library including Markov chain Monte Carlo (MCMC) and other methods.

3. A compiler that generates C++ for each model and algorithm, compiles the C++, and lets you use it from R.  You don't need to know anything about C++ to use nimble.

More information about NIMBLE can be found at [https://r-nimble.org](https://r-nimble.org).

The paper that describes NIMBLE is [here](https://www.tandfonline.com/doi/full/10.1080/10618600.2016.1172487).

### Getting help

The best way to seek user support is the nimble-users list.  Information on how to join can be found at [https://r-nimble.org](https://r-nimble.org).

### Illustrative Example:

For this vignette we will use the salamander example from the R package `glmmTMB`. Examples of this model can be seen in the general help file by running `?glmmTMB`. These data are repeated counts $y$ of salamanders at multiple sites, which are zero-inflated. Some of the sites are mined, and some are not($x$). Based on `glmmTMB`, we may analyse these data as

$$
  y_i|z_i \sim \text{Poisson}(z_i\lambda_i) 
$$

$$
  z_i \sim \text{Bernoulli}(1-p_i)
$$

$$
  \text{log}(\lambda_i) = \beta x_i + u_{site_i}
$$

$$
\text{logit}(p_i) = \alpha_0 x_i
$$

$$
u_{j} \sim \text{Normal}(0, \sigma^2)
$$

The maximum likelihood estimator can be found in `glmmTMB`.

```{r, message = FALSE, warning = FALSE}
library(glmmTMB)
data(Salamanders)
m1 <- glmmTMB(count ~ spp * mined + (1|site), zi=~1, family=poisson, data=Salamanders)
#summary(m1)
```

We can also match this model in `INLA` to get the approximate posterior as a "zeroinflatedpoisson1".

```{r, message = FALSE, warning = FALSE}
library(INLA)
fit.inla <- inla( count ~ spp * mined + f(site, model="iid"), 
  quantiles = c(.025,.25,.5,.75,.975), family= "zeroinflatedpoisson1", data=Salamanders, control.inla = list(int.strategy = "ccd"))
#summary(fit.inla)
```

## Building a GLMM in NIMBLE

To write this model in NIMBLE we will first write a zero-inflated Poisson distribution that marginalizes over the indirectly observed discrete value $z$. When $y=0$, $z$ is a discrete latent variable. In order to perform our methods, all the latent variables in this model must be continuous. `nimbleQuad` relies on automatic differentiation (AD) to take the derivatives. To include AD in a `nimbleFunction` we must set `buildDerivs = 'run'`. A random number generator is also provided for convenience, although this is not strictly needed for this model to fit.

```{r}
dZIP <- nimbleFunction(
 run = function(x = double(), lambda = double(),
                zeroProb = double(), log = logical(0, default = 0)) {
   returnType(double())
   ## For use with AD, we cannot use an `if` statement to handle the mixture.
   prob <- zeroProb * dbinom(x, size = 1, prob = 0) + (1 - zeroProb) * dpois(x, lambda)
   if (log) return(log(prob))
     return(prob)
   },
   buildDerivs = 'run'   # Needed when used with AD-based algorithms.
)

rZIP <- nimbleFunction(
 run = function(n = integer(), lambda = double(), zeroProb = double()) {
   returnType(double())
   isStructuralZero <- rbinom(1, prob = zeroProb, size = 1)
   if (isStructuralZero) return(0)
   return(rpois(1, lambda))
})
```

We will use prior distributions to match the `INLA` model that was fit above. See `inla.priors.used(fit.inla)` for a description of the default priors. INLA places a logit-normal prior on $p$. For convenience, we will write our own logit-normal prior. In this case, we explicitly register the distribution to tell NIMBLE the range values for parameter transformation to use. This is not a necessary step if not trying to match INLA.

```{r}
dlogitnormal = nimbleFunction(
  run = function(x = double(), mean = double(), sd = double(), 
      log = logical(0, default = 0)){
    ans <- dnorm(logit(x), mean = mean, sd = sd, log = TRUE) - log(x) - log(1-x)
    returnType(double())
    if(log) return(ans)
    else return(exp(ans))
  },
  buildDerivs = 'run'
)

registerDistributions(list(
  dlogitnormal = list(BUGSdist = "dlogitnormal(mean, sd)", range = c(0,1))
))
```

Following INLA, we use a completely flat prior on the intercept `dflat`, a gamma prior on precision `tau_re` and a logit-normal on `p`. We can now write the model code, and then provide constants and data to build the model. We are using AD and must set `buildDerivs = TRUE` when building the nimble model.

```{r}
## NIMBLE Model Code
code <- nimbleCode({
  beta[1] ~ dflat()
  for( i in 2:np ) 
    beta[i] ~ dnorm(0, tau = 0.001) 
  tau_re ~ dgamma(shape = 1, rate = 5e-5)
  p ~ dlogitnormal(mean = -1, sd = 1/sqrt(0.2))
  
  for(i in 1:nsites) 
    re[i] ~ dnorm(0, tau = tau_re)

  for( i in 1:nobs ){
    log(lam[i]) <- sum(beta[1:np]*X[i,1:np]) + re[site[i]]
    count[i] ~ dZIP(lambda = lam[i], zeroProb = p)
  }
})

nimconst <- list()
nimconst$X <- as.matrix(model.matrix( ~ spp * mined, data = Salamanders ))
nimconst$np <- ncol(nimconst$X)
nimconst$nobs <- nrow(nimconst$X)
nimconst$site <- as.numeric(factor(Salamanders$site))
nimconst$nsites <- max(nimconst$site)

nimdata <- list()
nimdata$count <- Salamanders$count

inits <- list(logitp = -0.9, 
  tau_re = 3,
  re = rnorm(nimconst$nsites,0,0.1),
  beta = rnorm(nimconst$np))

m <- nimbleModel(code, data = nimdata, constants = nimconst, inits = inits, buildDerivs = TRUE)
cm <- compileNimble(m)
```

## Using the Laplace Approximation

We can find the maximum likelihood estimate or posterior mode by using `buildLaplace` or `bulidAGHQ` if we want more quadrature nodes for marginalizing over the random effects. 

```{r}
laplace <- buildLaplace(model = m)
claplace <- compileNimble(laplace)

## Can find MLE manually:
mle <- claplace$findMLE()
summ.mle <- claplace$summary(mle, randomEffectsStdError = TRUE)

## Or use a wrapper:
fit.nimble <- runLaplace(claplace)

## Compare with glmmTMB:
params.nim <- fit.nimble$summary$params
fixef.nim <- params.nim[grep("beta", rownames(params.nim)),]

plot(fixef.nim[, "estimate"], fixef(m1)$cond, 
  xlab = "NIMBLE Estimate", ylab = "glmmTMB Estimate")
abline(a = 0, b = 1, lty = 2)

m1$fit$par["betazi"] - logit(params.nim["p",1])
exp(-2*m1$fit$par["theta"]) - params.nim["tau_re",1]
```

It can be helpful to check the Laplace approximation at the MLE by increasing the number of quadrature nodes using AGHQ. If the likelihood is very different when increasing the number of quadrature nodes then it indicates that Laplace was not a great approximation. In this particular case we have conditionally independent random effects so increasing the number of quadrature nodes will be relatively quick, but in other cases it can be very slow.

```{r}
claplace$calcLogLik(fit.nimble$summary$params[, 'estimate'], trans = FALSE)
claplace$updateSettings(nQuad = 5)
claplace$calcLogLik(fit.nimble$summary$params[, 'estimate'], trans = FALSE)
```


### Finding the Posterior Mode

Similar to `rstan`, we can use `optimize` within `buildAGHQ` or `buildLaplace` object to either penalize the likelihood, or find the posterior mode. We will assume that the user penalizes the likelihood by using prior distributions on parameters and those priors will be included in the likelihood calculations (`includePrior = TRUE`). If the posterior mode is of interest, then we must also include the Jacobian transformation by setting `includeJacobian = TRUE`. Finding the posterior mode, we can approximate the posterior distribution using the Hessian, similar to the MLE. In INLA, this is referred to as empirical Bayes (EB).

```{r}
claplace$updateSettings(nQuad = 1)
map <- claplace$optimize(includePrior = TRUE, includeJacobian = TRUE, parscale = "real")
summ.map <- summaryLaplace(claplace, map, randomEffectsStdError = TRUE, originalScale = TRUE)
summ.map$params
```

## Posterior Approximation

Posterior approximation is done by `buildNestedApprox`. Note that the setup is similar to `buildLaplace` except that we refer to the combination of random effects and fixed effects as `latentNodes` and the hyperparameters as `paramNodes`, which we will refer to more generally as the parameters. This reduces the dimension of the outer quadrature which reduces computation time. However, the fixed effects can also be made `paramNodes`. If nothing is put here NIMBLE will automatically detect these.

```{r}
approx <- buildNestedApprox(model = m, 
                            paramNodes = c('p', 'tau_re'), 
                            latentNodes = c('beta', 're'))
capprox <- compileNimble(approx, project = m)
```

The default posterior approximation methods using `runNestedApprox` are:

1. Find posterior mode of the parameters using Laplace to marginalize over the latent nodes.
2. Use an _integration free_ method analogous to INLA to approximate the marginal posterior distribution of each parameter.
3. Approximate the marginal log likelihood using the asymmetric multivariate normal approximation to the joint posterior of the parameters.
4. Calculate the posterior density of the parameters across the parameter quadrature grid (default = "CCD" when the number of parameters is > 2). A quadrature based marginal log likelihood is produced based on this computation.
5. Simulate the joint posterior of the `latentNodes` from a mixture of multivariate normal distributions with mean, covariance, and weights obtained from the parameter quadrature grid computation.

For each posterior density calculation of the parameters, Laplace is used to marginalize the latent nodes, which requires an inner optimization. The largest computational task is computing the posterior density on the parameter quadrature grid. The default of CCD, central composite design, is based on what INLA uses and scales well to higher dimensions. Alternatively, an AGHQ grid can be used which will be computationally more expensive but potentially more accurate. The user can also provide their own. By default, we do not do compute the parameter grid when calling `runNestedApprox` unless requested. It is necessary for inference on the latent nodes and it is automatically before the first call to `sampleLatents`.

Using the _integration free_ method is the fastest way to produce marginal posterior density estimates of the parameters. This is done by using an asymmetric multivariate normal distribution to approximate the posterior density of the parameters. Marginal distributions are easily computed from a fixed set of points, and then a spline is used to generate a density function. If more accuracy is required, quadrature can be used such as AGHQ to marginalize over the other parameters (similar to the R pacakge `aghq`). Here the user can choose between AGHQ, CCD, sparse AGHQ, or a user provided quadrature rule.

```{r}
result <- runNestedApprox(approx = capprox)
result

result$plotMarginal("tau_re")
lines(fit.inla$marginals.hyperpar[[2]], col = 'red', lty = 2)

result$plotMarginal("p")
lines(fit.inla$marginals.hyperpar[[1]], col = 'red', lty = 2)

result$sampleLatents(10000)
plot(density(result$samples[, "beta[1]"]), main = "", xlab = "beta[0]")
lines(fit.inla$marginals.fixed[[1]], col = 'red', lty = 2)
```

### Customizing the Quadrature Grid:

For the salamanders example, we have two parameters and default to using AGHQ as the parameter grid. The grid points are skewed based on the skew from the asymmetric multivariate normal approximation, and then scaled either using spectral decomposition (or Cholesky). The user may choose to not skew the points when computing the parameter grid. 

The user may also provide their own parameter grid. This must be written as a `nimbleFunction` with some specific components. A good choice for large dimension problems is the PCA-AGHQ rule (Howes et al. (2025) Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature). This rule uses AGHQ along the primary principal components of the spectral transformation and Laplace on the other dimensions. We can use the R package `mvQuad` to write our own simple version.

```{r}
RmvQuad <- function(levels, d) {
  out <- mvQuad::createNIGrid(dim=d, type = "GHe", level=levels)
  cbind(out$weights, out$nodes)
}
nimMVQuad <- nimbleRcall(function(levels = double(1), d = double()){}, Rfun = "RmvQuad", returnType = double(2))
PCA_AGHQ <- nimbleFunction(
    contains = QUAD_RULE_BASE,
    name = "quadRule_USER",
    setup = function() {},
    run = function() {},
    methods = list(
        buildGrid = function(levels = integer(0, default = 0), d = integer(0, default = 1)) {
            levs <- c(levels,1)
            output <- nimMVQuad(levels = levs, d = d)
            returnType(double(2))
            return(output)
        }
    )
)
```

We pass the user quadrature rule as a function in the control list `paramGridRule`. The default for a user supplied quadrature rule is that it constructs a multivariate grid (`paramGridRule_userType = "MULTI"`). If not, then the user function only returns a single dimension grid and then is turned into a grid via `"PRODUCT"` or `"SPARSE"`. Below we explicitly define the transformation method `quadTransform = "spectral"`, although this is the default.

```{r}
approx <- buildNestedApprox(model = m, 
                            paramNodes = c('p', 'tau_re'), 
                            latentNodes = c('beta', 're'), 
                            control = list(paramGridRule = PCA_AGHQ, 
                              paramGridRule_userType = "MULTI",
                              quadTransform = "spectral"))
capprox <- compileNimble(approx, project = m, resetFunctions = TRUE)

result <- runNestedApprox(approx = capprox)
```

We can look at the different quadrature grids by accessing the `capprox` object itself. Note that although `"AGHQSPARSE"` is available, it cannot be used to simulate the latent nodes as the quadrature weights can be  negative. For demonstration, we will compute the posterior density for each grid `calcParamGrid` to compare the different grids.

```{r}
capprox$buildParamGrid(quadRule = "USER")
nodes_user <- capprox$getParamGrid()
capprox$calcParamGrid()
marg_user <- capprox$calcMarginalLogLikQuad()

capprox$buildParamGrid(quadRule = "CCD")
nodes_ccd <- capprox$getParamGrid()
capprox$calcParamGrid()
marg_ccd <- capprox$calcMarginalLogLikQuad()

capprox$buildParamGrid(quadRule = "AGHQ")
nodes_aghq <- capprox$getParamGrid()
capprox$calcParamGrid()
marg_aghq <- capprox$calcMarginalLogLikQuad()

capprox$buildParamGrid(quadRule = "AGHQSPARSE")
nodes_aghqsparse <- capprox$getParamGrid()
capprox$calcParamGrid()
marg_aghqsparse <- capprox$calcMarginalLogLikQuad()

## Compare with multivariate asymmetric normal approximation
marg_sn <- capprox$calcMarginalLogLikApprox()
marg_aghq - marg_sn
marg_aghq - marg_ccd
marg_aghq - marg_user

## Visualize the grids
plot(nodes_user, xlim = c(-1.3, -0.6), ylim = c(0, 3), xlab = "theta[1]", ylab = "theta[2]")
points(nodes_aghq, pch = 4)
points(nodes_aghqsparse, pch = 3)
points(nodes_ccd, pch = 16)
legend("topleft", horiz=TRUE, legend = c("PCA-AGHQ", "AGHQ", "AGHQ-Sparse",  "CCD"), pch = c(1, 4, 3, 16), cex = 0.9)
```

An alternative to choosing a different grid is to use AGHQ and then prune a proportion of the grid, removing the points that contribute very little to the posterior density estimate. When running `buildNestedApprox`, we can control the number of quadrature points or the amount to prune the grid via `control = list(paramGridPrune = 0,  nQuadParam = 3)`.

The user should generally balance the amount of compute time it takes to run `calcParamGrid` with the amount of accuracy they want in estimation of the latent nodes. For a standard product construction of`nQuad`, with dimension `d`, we get `n^d` points on the quadrature grid. CCD does not scale as severely which is why it becomes recommended for higher dimensions. Alternatively, PCA-AGHQ could be applied, or heavy pruning of an AGHQ grid.

```{r}
## Increase the  number of AGHQ grid points:
capprox$buildParamGrid(quadRule = "AGHQ", nQuadUpdate = 11)
nodes_aghq11 <- capprox$getParamGrid()
capprox$calcParamGrid()
marg_aghq11 <- capprox$calcMarginalLogLikQuad()

## Prune Grid (Extreme)
capprox$buildParamGrid(quadRule = "AGHQ", nQuadUpdate = 11, prune = 0.5)
nodes_aghq11_50 <- capprox$getParamGrid()
marg_aghq11_50 <- capprox$calcMarginalLogLikQuad()
marg_aghq11 - marg_aghq11_50

plot(nodes_aghq11, pch = 4)
points(nodes_aghq11_50, col = 'red', pch = 1)
```

### Improving the Parameter Marginals

Although the _integration free_ method is fast and performs reasonably well, it may be that the user wishes to get a more exact marginal. This is done for each parameter of interest, marginalizing over the remaining parameters using an outer quadrature rule (AGHQ, CCD, AGHQSPARSE, USER). We choose the number of quadrature points for marginalization `nQuad` as well as the number of points along a grid to evaluate the marginal `nMarginalGrid`. 

The user may also choose to simulate from the asymmetric multivariate normal approximation to make inference on the marginals of the parameters. This can be particularly useful when making inference on parameters with dimension changes (e.g. proportions with Dirichlet prior).

```{r}
result$improveParamMarginals(nodes = 'tau_re', 
  nMarginalGrid = 9, nQuad = 3, quadRule = "AGHQ")

param.smp <- result$sampleParams(10000)
result$plotMarginal("tau_re")
lines(density(param.smp[,2]), col = 'blue', lty = 2)
lines(fit.inla$marginals.hyperpar[[2]], col = 'red', lty = 2)
```
